Functional traits complicate the picture of temporal biodiversity change in bird and mammal communities

### 2024-03-15 paper/GEB Submission/manuscript_GEB.tex
Look at study locations.

\title{Prevailing scenarios of functional change in Anthropocene bird
and mammal communities}

@@

Here we
\textcolor{blue}{leverage ongoing efforts to assemble functional trait data and computational advances to}
perform the first multi-taxa, multi-realm assessment of functional
diversity change through time. We focus on mammal and bird species as a
significant subset of the world's biodiversity heavily impacted by
anthropogenic change. While examining trends in plants, invertebrates,
and other vertebrate species is of equal interest, trait data for those
taxa raise additional challenges such as limited and biased species
coverage (35), a lack of accepted species-level means, and differences
in the types of traits collected. To ensure comparability across taxa in
trait type and data quality we therefore focus on mammals and birds.
Traits were intentionally selected to be representative of a species'
Eltonian niche, thereby summarizing the functional role they play in the
community (36). An initial assessment of amphibian trends is included in
the supplement, but excluded from general trend assessment here due to
limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series
to determine whether or not
\textcolor{blue}{the addition of functional trait data gives a clarifying picture of biodiversity change across communities. Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination. We present results at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm), and trends for individual studies.}

@@

Our final dataset included 2,432 time series from 50 studies in 21
countries and 12 biomes and 6 different traits (Fig \ref{fig:taxaMap}).
Data came from both terrestrial and marine realms and five climates
(Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical). The
earliest sample was in 1923 and the most recent was in 2014. For a full
breakdown of studies and their characteristics, see the supplement. Our
final dataset reflects many of the data biases that make global
synthesis work challenging, including geographic bias, a bias away from
areas currently under the greatest threat, and a bias towards shorter
time series. We address these shortcomings and their potential impact on
our results in the discussion.

@@

To test for trends within and between different levels of taxa, biome,
and realm we fit separate models with each of those factors added as a
predictor to the original model structure. We estimated within-level
slopes and calculated between-level contrasts using the \emph{emmeans}
package (v1.8.2, 47). We assessed the impact of time series duration and
start year on study-level trends using linear models with duration and
start year as predictors. All models in our analysis were fit using the
\emph{lme4} (v1.1-30) package in R (v4.2.3) and p-values were calculated
by Satterthwaite's degrees of freedom method using the \emph{lmerTest}
(v3.1-3) package with a significance level of \(\alpha = 0.05\)
(48--50).

@@

We found no significant overall trend in species richness or functional
diversity metrics (observed or \textcolor{blue}{corrected}) (Fig
\ref{fig:timeseriesPlot}). We did find a significant overall decrease in
Jaccard similarity, indicating accumulating changes in species
composition. Non-significant overall trends indicate that although some
studies experience increasing or decreasing trends, the average trend
across studies was plausibly 0 (Table \ref{tab:resultsTab}).
\textcolor{blue}{Within-group} trends for different taxa, biomes, or
realms were also non-significant for richness and functional diversity
metrics, with the exception of a significantly increasing trend for
functional evenness of global studies (characterized by having samples
on multiple continents), and a significantly decreasing functional
richness slope for temperate/tropical studies and mammal studies.
However, trends were not exhibited for the \textcolor{blue}{corrected}
metric indicating that differences
\textcolor{blue}{in functional diversity metrics} were largely due to
changes in species richness. Further, with only two global studies, the
trend should not be considered truly general. The general trends for
\emph{CWM} models were similarly not significant, with a significant
positive trend for only percentage of fish in diet composition.

@@

We did find significant differences between taxa, realms, and climate
for Jaccard similarity and some of the \emph{CWM's}. For example, while
Jaccard similarity was decreasing in the general trend and there were
significant within group slopes for Mammals, Birds, and Terrestrial
communities, there was no significant slope for the marine realm,
indicating that the general trend is mostly driven by turnover in
terrestrial communities. For bird communities, we also found within
group trends and between group differences for trends in foraging
behavior. We found a significant increasing trend in utilization of the
canopy in Tropical communities that was significantly different from the
trends for Polar/Temperate and Temperate communities. There was a
significant decrease in utilization of the understory for Terrestrial
communities and significant increase in foraging below the water surface
for global studies (but see the previous limitations of Global data).
There was also a significant positive slope for \emph{CWM} body mass for
the single Temperate/Tropical study (with three time series) which were
marine mammal communities.

We found significant dietary shifts across communities, with a
significant increase in fruit consumption in Terrestrial communities and
a significant decrease in nectar consumption in Tropical communities, a
trend significantly different than that for Terrestrial communities.
There was a significant increase in seed consumption in bird species,
which was significantly different from the trend for Mammal communities.
There was a significant increase in fish consumption for the two Global
studies. Vertebrate consumption significantly decreased for Marine
studies and for studies of Global and Tropical communities.

@@

Our study represents the largest broad-scale multi-taxa assessment of
functional change through time to date, giving a first look at aggregate
and local trends in functional diversity in mammal and bird communities.
\textcolor{blue}{Our results show that the addition of functional traits illuminates a few consistent functional trends across communities, but largely complicates rather that clarifies the story of biodiversity change. While the characteristics of species clearly matter, instead of unifying the nature of communities' change, they more often distinguish them. In general, we found a few areas of consensus for models where communities were aggregated (general trends, trends by taxa, biome, and realm), and vast heterogeneity for study-level models.}

@@

We did not detect an overall trend in any functional diversity metrics.
As with previous species-based syntheses, we also found no overall trend
in species richness accompanied by increasing turnover through time
(31), indicating that non-significant trends in functional metrics may
be consistent with similar well-documented species derived trends. We
found no trend in functional change for almost all realms, biomes, and
taxonomic groups.
\textcolor{blue}{The one exception was a consensus trend indicating}
functional richness loss for mammal studies. These results are
consistent with multiple studies linking anthropogenic drivers to loss
of functional diversity in mammal communities (51--53).

Despite a lack of general trends in functional diversity metrics, we did
find evidence of compositional shifts in the trait space with multiple
\texttt{significant} trends in \emph{CWM}'s for studies coming from the
same taxa, realms, or climates,
\texttt{indicating\ areas\ of\ consensus\ change\ across\ communities}.
Some of those trends are consistent with existing work in those
contexts. For example, the decrease in insect consumption for tropical
communities reflects well documented declines in tropical insectivorous
birds (54). Reduction in utilization of understory foraging for
terrestrial communities could be the result of disturbance like human
recreation or increased predation from introduced species. Other shifts
in diet for both Birds and Mammals, including increasing fruit, seed,
and vertebrate consumption and decreasing seed consumption in some
climates point to important areas for further exploration.

@@

The strongest picture the study-level results give is that trait level
data further complicates the story of biodiversity change. The lack of
general trend in functional diversity metrics belies a huge range of
positive and negative trends at the study-level that happen to average
out to zero. In order to simplify discussion, we will talk about the
implications for communities grouped by their concurrent change in
functional diversity metrics, species richness and turnover split into
positive, negative or no trend. While there are over 150 possible
combinations of change direction (or no change) in the 5 metrics, we
discuss here the six scenarios that occurred in more than one study: no
change, species loss or gain only, loss of functional evenness, richness
loss with species turnover, and increase in species and functional
richness accompanied by significant turnover. We focus on these
scenarios to combat potential spurious results due to multiple testing,
as it is unlikely all observations of the same scenario are spurious.
The number of studies in each scenarios is broken down in table x to
illustrate the relative number between groups rather than absolute
number.

The majority group of studies exhibited no trend in any species or
functional diversity metric.

At the study level, biodiversity change fell into four dominant
scenarios of change based on trends in richness, turnover, and
functional metrics: no change, only functional change, richness and
functional change with turnover, and richness and functional change
without turnover (Table \ref{tab:changeScenarios}). In contrast to our
expectation based on anthropogenic impacts, around a quarter of studies
experienced no trend in richness, turnover, or functional metrics. These
studies spanned the distribution of study durations excluding only the
very longest running studies, with the longest no change time series
lasting 23 years.

Another quarter of the studies showed no trend in richness or turnover,
but did show a significant shift in at least one summary functional
metric. These communities are best aligned with scenarios of species
replacement where only a few functional outliers replaced functionally
indistinct species or vice versa (there were studies with positive and
negative trends in all three functional metrics). Studies in the
scenario of only functional change were heavily skewed toward shorter
running time series, indicating that they may represent a limited
snapshot of communities that on longer time scales would be exhibiting
significant change in species-based metrics, particularly as species
replacement continued adding to turnover. Communities in this scenario,
and the previous scenario of no functional change are consistent with
hypotheses that species richness may be strongly regulated through time
(3, 8), and emphasize that maintenance of functional structure can be
divorced from those processes.

About 10 percent of studies, but the majority of time series in our
dataset, exhibited richness and functional change but no turnover. These
studies also spanned the the distribution of time series durations,
excluding the longest running studies. With the exception of one study,
trends in functional change for this group were only found for
uncorrected metrics and disappeared after correcting for species
richness, indicating that functional change in this group is almost
exclusively the result of species gains or losses rather than turnover
of species with different functional traits.

About 15 percent of studies, which were also exclusively the longest
running studies, had a trend in turnover, richness, and functional
change. The studies in this scenario fell into two groups: a positive
trend in richness with a positive trend in functional richness
(corrected and uncorrected), and a negative trend in species richness
with a negative trend in uncorrected functional richness and no trend in
corrected richness but a trend in other corrected functional metrics.
\textcolor{blue}{This would indicate that when species are gained in a community, they disproportionately add to the functional richness. When species are lost however, they are not more functionally unique than expected by chance though they may produce shifts in other aspects of functional space.}

Our results illustrate the range of relationships between species-based
and functional metrics observed in real communities. Critically,
particularly for the short observation time windows that characterize
the majority of ecological data, functional shifts may be underlying
seemingly static species-based community metrics. Species-based
approaches may therefore be missing areas where significant change in
the functional space could be impacting ecosystem function and
resilience. In scenarios where there are significant species gains or
losses over time, we found evidence that species losses may be
relatively functionally indistinct, while species gains contribute
disproportionately to the richness of a functional space. This finding
is consistent with evidence that at least in bird communities, common
functionally general species are being lost even as rare species are
increasing (55--57).

Our results are consistent with ecological expectations that community
dissimilarity would be greater the larger the observation time window
purely due to background processes, which is likely further accelerated
by global change drivers (13). While we do not address how observed
rates differ from background expectation here, we can make some
assertions about the implication of turnover for functional change.
Where turnover is occurring, it is almost always accompanied by a
\textcolor{blue}{trend in one of the functional diversity metrics, indicating}
directional change in the functional space. Species are not therefore
being replaced by exact functional counter parts. The fact that we
observed both increases and decreases in corrected functional richness
accompanying turnover is a reminder that
\textcolor{blue}{new species are not always replacing more functionally unique species, but are often more functionally unique themselves.}

\begin{figure}
\includegraphics[width=\textwidth]{../../figures/duration_hist} \includegraphics[width=\textwidth]{../../figures/functional_change_table} \caption{Comparison of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange.}\label{fig:changeScenarios}
\end{figure}

Still, some results stand in contrast to predictions for trait shifts
under global change. For example, mean body size is predicted to
decrease as a result of climate change impacts and megafaunal loss (58),
a phenomenon which has already been well documented empirically and
experimentally in multiple taxa (59--62). While the species-level trait
means used here are not appropriate for assessing intraspecific body
size shifts, we would expect to see shifts in community-wide means due
to local losses of large-bodied species. Instead, we found no evidence
of a trend in \emph{CWM} body size, with the exception of a single study
of marine mammals in the Bahamas which showed a significant increase.
Still, many of the studies in our dataset draw from areas that may have
experienced significant loss of large-bodied species before the
observation window, with contemporary loss rates slowing (63). Trends
could be significantly different for the same time periods in regions of
sub-Saharan Africa, for example, which has poor representation in our
dataset but where megafauna exist on the landscape and are increasingly
threatened (64).

What does local maintenance of functional structure mean for ecosystem
function? The vast majority of experimental and observational work links
declines in function to declines in functional or species diversity (7,
65, 66). By those criteria very few communities in our dataset are in a
state of concern for loss of functionality (Table \ref{tab:trendTab}),
though we did find evidence that mammal communities may be in greater
danger than bird communities. Still, shifts in metrics are only relevant
if the underlying traits are those most critical for ecosystem function.
We were limited in this analysis to the traits available rather than
those with strong empirical links to function. Similarly, the dimensions
of functional space most important for ecosystem function are still a
topic of on going debate, and at least some known aspects important for
multifunctionality were not measured here (e.g.~dispersion, rarity,
abundance of dominant species, 67).

@@

While we found no over all trends in functional metrics, our results
should not be interpreted as an indication that the ongoing biodiversity
crisis is less severe than previously described, or that there is no
concern for functional change as a result of anthropogenic impact. In
fact, study-level trends indicate quite the opposite, that functional
shifts with unknown implications for ecosystem processes may be going
undetected by common species-based approaches, particularly for short
observation time windows. While the majority of studies in our dataset
did not experience a significant functional loss, a substantial body of
work links functional degradation to species losses as a result of
direct human intervention in the form of land use change and
intensification or habitat fragmentation, indicating that those studies
are simply not representative of the kinds of impacts of greatest policy
concern (26, 68, 69).

@@

We present here four prevailing scenarios of change experienced in bird
and mammal communities. While we offer a first discussion of which kinds
of change are most common and why, assessing the true global prevalence
will require continued efforts to fill data gaps, a well recognized
challenge in ecology (6, 12, 70). Still, a ``scenarios of change''
framework can provide structure for future work addressing functional
shifts, particularly as we reconcile results from broad-scale syntheses
with in-depth single system studies. It will be particularly critical to
link forms of change to individual drivers to assess which drivers may
impact species and functional diversity differently. Understanding those
links will help identify where directly measuring functional structure
instead of just species change is necessary for understand impacts on a
given system. In addition, direct links to drivers will improve our
ability to distinguish between communities experiencing no change due to
a lack of perturbation from communities with high resilience in the face
of disturbance.

Here we identify trends that are statistically significant, however they
may not necessarily be ecologically significant. While it is common to
link changes in functional metrics to changes in ecosystem processes,
those changes are less frequently discussed in terms of the size
necessary for ecological impact. The degree of change in a process
considered ecological meaningful is somewhat subjective and a function
of the system and management context, making identifying the ecological
meaning of broad-scale aggregate shifts even more challenging. It is
further hampered by the use of species-level trait data, as the traits
most closely linked to a process of interest may not be available (71).
Future trait collection that explicitly considers existing frameworks
for linking traits to processes (e.g.~the response and effect framework
15) would facilitate improved ecological interpretation of potential
functional changes.

@@

\leavevmode\vadjust pre{\hypertarget{ref-inger2015}{}}%
\CSLLeftMargin{55. }%
\CSLRightInline{Inger R, et al. (2015)
\href{https://doi.org/10.1111/ele.12387}{Common European birds are
declining rapidly while less abundant species' numbers are rising}.
\emph{Ecology Letters} 18(1):28--36.}

\leavevmode\vadjust pre{\hypertarget{ref-rosenberg2019}{}}%
\CSLLeftMargin{56. }%
\CSLRightInline{Rosenberg KV, et al. (2019) Decline of the North
American avifauna. \emph{Science}. Available at:
\url{https://www.science.org/doi/abs/10.1126/science.aaw1313}.}

\leavevmode\vadjust pre{\hypertarget{ref-schipper2016}{}}%
\CSLLeftMargin{57. }%
\CSLRightInline{Schipper AM, et al. (2016)
\href{https://doi.org/10.1111/gcb.13292}{Contrasting changes in the
abundance and diversity of North American bird assemblages from 1971 to
2010}. \emph{Global Change Biology} 22(12):3948--3959.}

\leavevmode\vadjust pre{\hypertarget{ref-sheridan2011}{}}%
\CSLLeftMargin{58. }%
\CSLRightInline{Sheridan JA, Bickford D (2011)
\href{https://doi.org/10.1038/nclimate1259}{Shrinking body size as an
ecological response to climate change}. \emph{Nature Climate Change}
1(8):401--406.}

\leavevmode\vadjust pre{\hypertarget{ref-caruso2014}{}}%
\CSLLeftMargin{59. }%
\CSLRightInline{Caruso NM, Sears MW, Adams DC, Lips KR (2014)
\href{https://doi.org/10.1111/gcb.12550}{Widespread rapid reductions in
body size of adult salamanders in response to climate change}.
\emph{Global Change Biology} 20(6):1751--1759.}

\leavevmode\vadjust pre{\hypertarget{ref-forster2012}{}}%
\CSLLeftMargin{60. }%
\CSLRightInline{Forster J, Hirst AG, Atkinson D (2012)
\href{https://doi.org/10.1073/pnas.1210460109}{Warming-induced
reductions in body size are greater in aquatic than terrestrial
species}. \emph{Proceedings of the National Academy of Sciences}
109(47):19310--19314.}

\leavevmode\vadjust pre{\hypertarget{ref-huss2019}{}}%
\CSLLeftMargin{61. }%
\CSLRightInline{Huss M, Lindmark M, Jacobson P, van Dorst RM, Gårdmark A
(2019) \href{https://doi.org/10.1111/gcb.14637}{Experimental evidence of
gradual size-dependent shifts in body size and growth of fish in
response to warming}. \emph{Global Change Biology} 25(7):2285--2295.}

>>>>

\title{Functional traits complicate the picture of temporal biodiversity
change in bird and mammal communities}

@@

Here we
\textcolor{blue}{leverage ongoing efforts to assemble functional trait data and recent computational advances to}
perform the first multi-taxa, multi-realm assessment of functional
diversity change through time. We focus on mammal and bird species as a
significant subset of the world's biodiversity heavily impacted by
anthropogenic change. While examining trends in plants, invertebrates,
and other vertebrate species is of equal interest, trait data for those
taxa raise additional challenges such as limited and biased species
coverage (35), a lack of accepted species-level means, and differences
in the types of traits collected. To ensure comparability across taxa in
trait type and data quality we therefore focus on mammals and birds.
Traits were intentionally selected to be representative of a species'
Eltonian niche, thereby summarizing the functional role they play in the
community (36). An initial assessment of amphibian trends is included in
the supplement, but excluded from general trend assessment here due to
limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series
to determine whether or not
\textcolor{blue}{the addition of functional trait data gives a clarifying picture of biodiversity change across communities. Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination. We present results at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.}

@@

Our final dataset included 2,432 time series from 50 studies in 21
countries and 12 biomes and 6 different traits (Fig \ref{fig:taxaMap}).
Data came from both terrestrial and marine realms and five climates
(Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical). The
earliest sample was in 1923 and the most recent was in 2014.
\textcolor{blue}{While it is not possible with this data to directly assess the level of human impact occurring for each study, we include binary protection status as a coarse indicator of impact level. However, protected areas were almost exclusively from temperate terrestrial studies (with one tropical study), so results are confounded by multiple other study characteristics.}
For a full breakdown of studies and their characteristics, see the
supplement. Our final dataset reflects many of the data biases that make
global synthesis work challenging, including geographic bias, a bias
away from areas currently under the greatest threat, and a bias towards
shorter time series. We address these shortcomings and their potential
impact on our results in the discussion.

@@

To test for trends within and between different levels of taxa, biome,
realm, and protection status we fit separate models with each of those
factors added as a predictor interacting with time to the original model
structure. We estimated within-level slopes and calculated between-level
contrasts using the \emph{emmeans} package (v1.8.2, 47). We assessed the
impact of time series duration and start year on study-level trends
using linear models with duration and start year as predictors. All
models in our analysis were fit using the \emph{lme4} (v1.1-30) package
in R (v4.2.3) and p-values were calculated by Satterthwaite's degrees of
freedom method using the \emph{lmerTest} (v3.1-3) package with a
significance level of \(\alpha = 0.05\) (48--50).

@@

We found no significant overall trend in species richness or functional
diversity metrics (observed or \textcolor{blue}{corrected}) (Fig
\ref{fig:timeseriesPlot}). We did find a significant overall decrease in
Jaccard similarity, indicating accumulating changes in species
composition. Non-significant overall trends indicate that although some
studies experience increasing or decreasing trends, the average trend
across studies was plausibly 0 (Table \ref{tab:resultsTab}).
\textcolor{blue}{Within-group} trends for different taxa, biomes,
realms, or protection statuses were also non-significant for richness
and functional diversity metrics, with the exception of a significantly
increasing trend for functional evenness of global studies
(characterized by having samples on multiple continents), and a
significantly decreasing functional richness slope for
temperate/tropical studies and mammal studies. However, trends were not
exhibited for the \textcolor{blue}{corrected} metric indicating that
differences \textcolor{blue}{in functional diversity metrics} were
largely due to changes in species richness. Further, with only two
global studies, the trend should not be considered truly general. The
general trends for \emph{CWM} models were similarly not significant,
with a significant positive trend for only percentage of fish in diet
composition.

@@

We did find significant differences between taxa, realms, and climate
for Jaccard similarity and some of the \emph{CWM's}. For example, while
Jaccard similarity was decreasing in the general trend and there were
significant within group slopes for Mammals, Birds, and Terrestrial
communities, there was no significant slope for the marine realm,
indicating that the general trend is mostly driven by turnover in
terrestrial communities. We also found a significant decrease in Jaccard
similarity in unprotected areas only, with no trend for protected areas.
For bird communities, we also found within group trends and between
group differences for trends in foraging behavior. We found a
significant increasing trend in utilization of the canopy in Tropical
communities that was significantly different from the trends for
Polar/Temperate and Temperate communities. There was a significant
decrease in utilization of the understory for Terrestrial communities
and significant increase in foraging below the water surface for global
studies (but see the previous limitations of Global data). There was
also a significant positive slope for \emph{CWM} body mass for the
single Temperate/Tropical study (with three time series) which were
marine mammal communities.

We found significant dietary shifts across communities, with a
significant increase in fruit consumption in Terrestrial communities and
a significant decrease in nectar consumption in Tropical communities, a
trend significantly different than that for Terrestrial communities.
There was a significant increase in seed consumption in bird species,
which was significantly different from the trend for Mammal communities.
There was a significant increase in fish consumption for the two Global
studies. Vertebrate consumption significantly decreased for Marine
studies and for studies of Global and Tropical communities.
\textcolor{blue}{There was also a significant increase in fish and fruit consumption for unprotected areas, but not protected.}

@@

Our study represents the largest broad-scale multi-taxa assessment of
functional change through time to date, giving a first look at aggregate
and local trends in functional diversity in mammal and bird communities.
\textcolor{blue}{Our results show that the addition of functional traits illuminates a few consistent functional trends across communities, but largely complicates rather that clarifies the story of biodiversity change. While the characteristics of species clearly matter, instead of unifying the nature of communities' change, they more often distinguish them. In general, we found a few areas of consensus for models where communities were aggregated (general trends, trends by taxa, biome, realm, protection status), and vast heterogeneity for study-level models.}

@@

\textcolor{blue}{The most stark area of consensus was the lack of general trend aggregating across communities.}
We did not detect an overall trend in any functional diversity metrics.
As with previous species-based syntheses, we also found no overall trend
in species richness accompanied by increasing turnover through time
(31), indicating that non-significant trends in functional metrics are
consistent with similar well-documented species derived trends. We found
no trend in functional change for almost all realms, biomes, and
taxonomic groups.
\textcolor{blue}{The one exception was a consensus trend indicating}
functional richness loss for mammal studies. These results are
consistent with multiple studies linking anthropogenic drivers to loss
of functional diversity in mammal communities (51--53).

Despite a lack of general trends in functional diversity metrics, we did
find evidence of compositional shifts in the trait space with multiple
\texttt{significant} trends in \emph{CWM}'s for studies coming from the
same taxa, realms, climates, or protection statuses,
\texttt{indicating\ areas\ of\ consensus\ change\ across\ communities}.
Some of those trends are consistent with existing work in those
contexts. For example, the decrease in insect consumption for tropical
communities reflects well documented declines in tropical insectivorous
birds (54). Reduction in utilization of understory foraging for
terrestrial communities could be the result of disturbance like human
recreation or increased predation from introduced species. Other shifts
in diet for both Birds and Mammals, including increasing fruit, seed,
and vertebrate consumption and decreasing seed consumption in some
climates point to important areas for further exploration.

@@

\textcolor{blue}{At the study-level, our results run contrary to our hope for a clearer picture of biodiversity change through a functional lens. The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level. In order to simplify discussion, we will talk about the implications for ecosystem process and vulnerability for communities grouped by their concurrent change in functional diversity metrics, species richness and turnover split into positive, negative or no trend. While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover. We focus on these scenarios to combat potential spurious results due to multiple testing, as it is unlikely all observations of the same scenario are due to false positives. We include a break down of the number of studies in each scenario to illustrate relative number of studies in each group, but emphasize that the absolute number is likely impacted by multiple testing.}

The majority group of studies exhibited no trend in any species or
functional diversity metric. Contrary to the expectation due to
anthropogenic and global change stressors, these communities do not show
significant changes over the course of the observation window. Studies
in this group span the distribution of study durations excluding only
the very longest running studies, with the longest no change time series
lasting 23 years. They also included both bird and mammal studies and
only four were located in protected areas, indicating that the lack of
trend is not restricted to a specific ecological context or those
communities most insulated from human impact.

The lack of trend could be the result of multiple possible scenarios.
First, these may be communities resisting perturbations or simply not
experiencing significant perturbations. Given the studies in this group
come from all possible taxa, realms, climates, and protection statuses,
evidence points to communities resisting perturbation, offering some
hope that there are areas of the globe where communities are fairing
well for now. Alternatively, these may be communities that have
experienced or continue to experience significant stress, but lost
species or functional diversity outside the observation window. This
could be true particularly for North American mammal communities where
trophic downgrading and megafaunal losses occurred hundreds of years ago
{[}citation{]}. Third, these communities may be experiencing directional
shifts undetectable by available data. For example, species-level trait
data does not capture intraspecific shifts in the trait space, which can
represent significant changes in overall functional composition and
impact maintenance of ecological processes {[}citation{]}.

Three of the groups fit under the broad umbrella of changes in
redundancy. By definition, if a community gains or loses species while
functional metrics are unchanged, those species represent an increase or
decrease in redundancy, so we include communities exhibiting gains or
losses in species richness with no change in functional metrics or
turnover and richness loss with significant turnover in this umbrella.
For communities exhibiting loss of redundancy (declining species
richness), ecological processes are likely being maintained, but
capacity to respond to future stressors is reduced (55). These
communities are actually fairing better than expected looking at
species-based metrics alone, but are also in a precarious position for
maintaining ecological function into the future (56). Conversely,
communities exhibiting a increase in redundancy (via an increase in
species richness) are becoming better positioned to respond to future
stressors, but are not actually expanding their functional capacity as
may be assumed by looking at species gains alone (57).

The next group of studies exhibit an increase in species richness and
functional richness with significant turnover. While these communities
are losing some species that are functionally redundant or functional
analogs of species additions, they are gaining even more species that
expand the functional space. These communities have a hopeful
trajectory, as they are improving their functional capacity and
potentially their functional redundancy, likely leading to robust future
ecological function. Notably, the studies that exhibited species
richness increases (in this group with increasing functional richness,
as well as communities without functional richness increases) were
exclusively from terrestrial, temperate (sometimes temperate/polar) bird
communities and occurred in countries where there has been a significant
investment in conservation over the last few decades (United States,
Canada, Sweden). Our results are consistent with other functional work
in these regions, showing increases in species richness and functional
diversity for North American breeding birds over the last few decades
after a period of decline(58), and loss of common, functionally general
species even as rare species are increasing in North America and Europe
(59--61).

The final group exhibits decreasing functional evenness. The empirical
link between changes in evenness and ecosystem process is the most
poorly studied of the metrics, however theory indicates that increased
evenness improves function and stability (62). If one or a few species
become largely dominant in a community, function will be the result of
only those species rather than spread out across a wealth of species
responding independently to environmental change (42, 63). Further,
there is some evidence that maintaining evenness is important for
supporting multifunctionality in communities (64). These communities are
therefore of particular concern, especially considering evidence
evenness is more sensitive to environmental change that richness (62).

While the majority of studies fell into one of the groups discussed in
detail above, 25\% of the studies exhibited some different combination
of change in metrics, underlining the vast heterogeneity of realized
change scenarios. We have an indication of the ecological
characteristics most likely to lead to some of the change types
(e.g.~temperate bird communities are gaining species and functional
richness), but most types of change are exhibited across many different
kinds of communities. Identifying key factors mediating the kind of
biodiversity change a community experiences is critical for identifying
areas of conservation concern and extrapolating results to areas where
functional trait data may not be available. While it is tempting to draw
a one to one line between the degree of human impact and loss of
functional integrity (in the form of functional redundancy and
richness), that relationship is not borne out by our rough metric of
human impact or previous taxa-specific or single community work, which
is largely mixed (26, 65--67).

\hypertarget{areas-of-incongruence}{%
\subsubsection{Areas of incongruence}\label{areas-of-incongruence}}

\textcolor{blue}{The addition of functional trait illuminated a few other key areas where a synthesis approach is incongruous with system-specific studies or expectations based on anthropegenic impacts.}
For example, mean body size is predicted to decrease as a result of
climate change impacts and megafaunal loss (68), a phenomenon which has
already been well documented empirically and experimentally in multiple
taxa (69--72). However, we found no evidence of a trend in \emph{CWM}
body size, with the exception of a single study of marine mammals in the
Bahamas which showed a significant increase.
\textcolor{blue}{While the lack of trend could be explained at least in part by data limitations such as shortened observation windows or missing intraspecific data, the fact that only one of the 41 studies for which study-level models were fit showed a change in mean body size, and in the opposite direction of the expectation, is suprising and warrants further reconciliation.}

Results for protected areas were also surprising given

\begin{itemize}
\item
  changes in diet
\item
  turnover outside protected areas, not inside ( though might be driven
  by the fact that there's generally less turnover in terrestrial and
  temperate studies) (37)
\end{itemize}

@@

While we found no over all trends in functional metrics, our results
should not be interpreted as an indication that the ongoing biodiversity
crisis is less severe than previously described, or that there is no
concern for functional change as a result of anthropogenic impact. In
fact, study-level trends indicate quite the opposite, that functional
shifts with unknown implications for ecosystem processes may be going
undetected by common species-based approaches, particularly for short
observation time windows. While the majority of studies in our dataset
did not experience a significant functional loss, a substantial body of
work links functional degradation to species losses as a result of
direct human intervention in the form of land use change and
intensification or habitat fragmentation, indicating that those studies
are simply not representative of the kinds of impacts of greatest policy
concern (67).

@@

We present here four prevailing scenarios of change experienced in bird
and mammal communities. While we offer a first discussion of which kinds
of change are most common and why, assessing the true global prevalence
will require continued efforts to fill data gaps, a well recognized
challenge in ecology (6, 12, 73). Still, a ``scenarios of change''
framework can provide structure for future work addressing functional
shifts, particularly as we reconcile results from broad-scale syntheses
with in-depth single system studies. It will be particularly critical to
link forms of change to individual drivers to assess which drivers may
impact species and functional diversity differently. Understanding those
links will help identify where directly measuring functional structure
instead of just species change is necessary for understand impacts on a
given system. In addition, direct links to drivers will improve our
ability to distinguish between communities experiencing no change due to
a lack of perturbation from communities with high resilience in the face
of disturbance.

Here we identify trends that are statistically significant, however they
may not necessarily be ecologically significant. While it is common to
link changes in functional metrics to changes in ecosystem processes,
those changes are less frequently discussed in terms of the size
necessary for ecological impact. The degree of change in a process
considered ecological meaningful is somewhat subjective and a function
of the system and management context, making identifying the ecological
meaning of broad-scale aggregate shifts even more challenging. It is
further hampered by the use of species-level trait data, as the traits
most closely linked to a process of interest may not be available (74).
Future trait collection that explicitly considers existing frameworks
for linking traits to processes (e.g.~the response and effect framework
15) would facilitate improved ecological interpretation of potential
functional changes.

@@

\leavevmode\vadjust pre{\hypertarget{ref-reich2012}{}}%
\CSLLeftMargin{55. }%
\CSLRightInline{Reich PB, et al. (2012)
\href{https://doi.org/10.1126/science.1217909}{Impacts of biodiversity
loss escalate through time as redundancy fades}. \emph{Science (New
York, NY)} 336(6081):589--592.}

\leavevmode\vadjust pre{\hypertarget{ref-pimiento2020a}{}}%
\CSLLeftMargin{56. }%
\CSLRightInline{Pimiento C, et al. (2020)
\href{https://doi.org/10.1098/rspb.2020.1162}{Selective extinction
against redundant species buffers functional diversity}.
\emph{Proceedings of the Royal Society B: Biological Sciences}
287(1931):20201162.}

\leavevmode\vadjust pre{\hypertarget{ref-biggs2020}{}}%
\CSLLeftMargin{57. }%
\CSLRightInline{Biggs CR, et al. (2020)
\href{https://doi.org/10.1002/ecs2.3184}{Does functional redundancy
affect ecological stability and resilience? A review and meta-analysis}.
\emph{Ecosphere} 11(7):e03184.}

\leavevmode\vadjust pre{\hypertarget{ref-jarzyna2016}{}}%
\CSLLeftMargin{58. }%
\CSLRightInline{Jarzyna MA, Jetz W (2016)
\href{https://doi.org/10.1111/gcb.13571}{A near half-century of temporal
change in different facets of avian diversity}. \emph{Global Change
Biology} 23(8):2999--3011.}

\leavevmode\vadjust pre{\hypertarget{ref-inger2015}{}}%
\CSLLeftMargin{59. }%
\CSLRightInline{Inger R, et al. (2015)
\href{https://doi.org/10.1111/ele.12387}{Common European birds are
declining rapidly while less abundant species' numbers are rising}.
\emph{Ecology Letters} 18(1):28--36.}

\leavevmode\vadjust pre{\hypertarget{ref-rosenberg2019}{}}%
\CSLLeftMargin{60. }%
\CSLRightInline{Rosenberg KV, et al. (2019) Decline of the North
American avifauna. \emph{Science}. Available at:
\url{https://www.science.org/doi/abs/10.1126/science.aaw1313}.}

\leavevmode\vadjust pre{\hypertarget{ref-schipper2016}{}}%
\CSLLeftMargin{61. }%
\CSLRightInline{Schipper AM, et al. (2016)
\href{https://doi.org/10.1111/gcb.13292}{Contrasting changes in the
abundance and diversity of North American bird assemblages from 1971 to
2010}. \emph{Global Change Biology} 22(12):3948--3959.}

\leavevmode\vadjust pre{\hypertarget{ref-filstrup2019}{}}%
\CSLLeftMargin{62. }%
\CSLRightInline{Filstrup CT, King KBS, McCullough IM (2019)
\href{https://doi.org/10.1111/ele.13407}{Evenness effects mask richness
effects on ecosystem functioning at macro-scales in lakes}.
\emph{Ecology Letters} 22(12):2120--2129.}

\leavevmode\vadjust pre{\hypertarget{ref-hillebrand2008}{}}%
\CSLLeftMargin{63. }%
\CSLRightInline{Hillebrand H, Bennett DM, Cadotte MW (2008)
\href{https://doi.org/10.1890/07-1053.1}{Consequences of Dominance: A
Review of Evenness Effects on Local and Regional Ecosystem Processes}.
\emph{Ecology} 89(6):1510--1520.}
